---
title: "Melanoma Statistical Analysis in R"
output: html_document
---

```{r setup, include=FALSE}
library(boot)
library(dplyr)
library(ggplot2)
library(caret)
library(rpart.plot)
```

## Data preparation

The project uses `boot::melanoma`, so no external data file is required.

```{r}
set.seed(1821)

melanoma <- boot::melanoma %>%
  mutate(
    sex = factor(sex, levels = c(0, 1), labels = c("male", "female")),
    ulcer = factor(ulcer, levels = c(0, 1), labels = c("absent", "present")),
    # In boot::melanoma, status 1 denotes death from melanoma.
    melanoma_death = factor(
      if_else(status == 1, "melanoma_death", "other_or_alive"),
      levels = c("melanoma_death", "other_or_alive")
    )
  )

str(melanoma)
summary(melanoma)
```

## Descriptive analysis

```{r}
melanoma %>%
  count(ulcer) %>%
  mutate(proportion = n / sum(n))
```

```{r}
melanoma %>%
  group_by(ulcer) %>%
  summarise(
    mean_time = mean(time, na.rm = TRUE),
    median_time = median(time, na.rm = TRUE),
    sd_time = sd(time, na.rm = TRUE),
    iqr_time = IQR(time, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```

```{r}
ggplot(melanoma, aes(ulcer, time, fill = ulcer)) +
  geom_boxplot(show.legend = FALSE) +
  theme_minimal() +
  labs(
    title = "Survival/follow-up time by ulceration status",
    x = "Ulceration",
    y = "Time (days)"
  )
```

```{r}
ggplot(melanoma, aes(time, fill = ulcer)) +
  geom_histogram(bins = 18, alpha = 0.75, position = "identity") +
  facet_grid(sex ~ ulcer) +
  theme_minimal() +
  labs(title = "Time distribution by sex and ulceration status")
```

## Non-parametric comparison and correlation

```{r}
thickness_by_sex <- wilcox.test(
  thickness ~ sex,
  data = melanoma,
  alternative = "two.sided",
  conf.level = 0.95
)
thickness_by_sex

thickness_time_correlation <- cor(
  melanoma$thickness,
  melanoma$time,
  method = "spearman",
  use = "complete.obs"
)
thickness_time_correlation
```

```{r}
ggplot(melanoma, aes(time, thickness, color = ulcer)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE) +
  theme_minimal() +
  labs(title = "Tumour thickness and observed time")
```

## Stratified hold-out split

The original three-level status variable distinguishes melanoma death, survival and death from other causes. For this educational binary classification example, `melanoma_death` compares melanoma death with the combined `other_or_alive` group. That simplification is a limitation and should not be interpreted as a clinical endpoint definition.

```{r}
model_data <- melanoma %>%
  select(melanoma_death, age, year, thickness, ulcer, sex)

set.seed(1821)
train_index <- createDataPartition(
  model_data$melanoma_death,
  p = 0.75,
  list = FALSE
)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]
```

## Logistic regression

```{r}
control <- trainControl(
  method = "repeatedcv",
  number = 5,
  repeats = 5,
  classProbs = TRUE,
  savePredictions = "final"
)

set.seed(1821)
glm_model <- train(
  melanoma_death ~ .,
  data = train_data,
  method = "glm",
  family = binomial(),
  trControl = control,
  metric = "Accuracy"
)

coef(glm_model$finalModel)
summary(glm_model$finalModel)
```

## Held-out classification evaluation

```{r}
glm_predictions <- predict(glm_model, newdata = test_data)
glm_confusion <- confusionMatrix(glm_predictions, test_data$melanoma_death)
glm_confusion
```

## Regression tree for tumour thickness

This secondary example predicts tumour thickness only from variables available at or near baseline (`sex`, `age`, `year`, `ulcer`). Follow-up time and outcome status are deliberately excluded because they occur after the tumour measurement and would create temporal/post-outcome leakage in a predictive framing.

```{r}
regression_data <- melanoma %>%
  select(thickness, sex, age, year, ulcer)

set.seed(1821)
reg_index <- createDataPartition(
  cut(regression_data$thickness, breaks = 5),
  p = 0.75,
  list = FALSE
)
reg_train <- regression_data[reg_index, ]
reg_test <- regression_data[-reg_index, ]

set.seed(1821)
tree_model <- train(
  thickness ~ .,
  data = reg_train,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 10),
  tuneLength = 20
)

rpart.plot(tree_model$finalModel)
tree_predictions <- predict(tree_model, newdata = reg_test)
postResample(tree_predictions, reg_test$thickness)
```

## Scope

This is an academic statistical-programming exercise using a historical teaching dataset. It is **not** a clinical decision-support model and the simplified binary outcome should not be used for medical interpretation.

## Session information

```{r}
sessionInfo()
```
